### 2019-11-12 paper/conceptual_framework.tex
Add draft discussion of son preference and sex selection
Very much a draft at the moment.
Need to add sex selection theory predictions

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Molitoris2019}.

@@

Effect of son preference on birth spacing?
Do we have any theoretical results on this?

@@

Women with different education levels have different hazard profiles, although the size 
and direction of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Furthermore, the use of sex selection increases with education, either because of the 
associated lower fertility or because higher income enables them to access and
afford prenatal sex determination
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining son preference \citep{bhat03,pande07}.
}

@@

Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}

For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.
}
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.

@@

In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.

>>>>

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Whitworth2002,Molitoris2019}.

@@

Finally, the discussion has so far assumed that children are identical, except for any potential
differences that arise from differences in spacing, but there is substantial evidence that the 
sex composition of previous children affect birth spacing.
For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
There is also evidence of differential stopping behavior, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.
}

It is not immediately obvious that we should expect a result like this from a theoretical
standpoint.
If, as discussed above, there are potentially significant negative effects of short spacing
parents who choose short spacing in the absence of a boy lowers the survival probabilty of
the next child, including a boy if the next child is a boy.
There are two possibilities for why parents might still try to conceive sooner in the 
absence of boys.%
\footnote{
The fact that there are significant differences in spacing by sex composition indicates that
parents have---at least some---control over the timing of birth.
Hence, it is unlikely that the differences in spacing come about by random chance.
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
}
First, it is possible that parents do not, in fact, realize that short spacing is potentially
detrimental to the next child's health.
Second, parents realize that short spacing has negative health effects, but have a strong
enough son preference that they are willing to divert substantial resources to investing
in the next child's health if it is a boy and that these investments can compensate for 
the negative effects.
[Are there any other ones that might explain the pattern?
    - biological? this would have to work off the sex of only the last child, so that
    spacing after girls is naturally shorter.
    - parents care so much about having a son sooner that they are willing to accept a 
    higher mortality risk.
    - ??
    ]

[has anyone actual done the theory behind the short spacing?
It would seem that it requires fairly strong assumptions to make it attractive to
have short spacing.
Maybe check the Jensen paper on equal treatment]

[can I say anything about which one is true?]
It is worthwhile thinking a bit about under what circumstances it would be possible to
identify which of these explanations hold.
If the first one holds, we should see high mortality for both boys and girls born
after short spacing across all households whether or not they have a strong son preference.
In fact, it is possible that we see higher mortality for sons than girls given that boys 
have naturally higher mortality in the absense of any preference for a given sex.

If the second one holds, we should see substantial differences in mortality across
sex for short spacing children, with the boys having substantially lower mortality risk
than expected for the spacing length.

If a son is born next:
The boy should have substantially lower mortality risk than expected for the spacing
duration.
Because this argument relies on a redistribution of resources, we should expect to see
worse outcomes for all the other children in the family.
This means that the other children (all girls) should do substantially worse.

If a girl is born next:
there are a couple of possible outcomes.
If all children/girls are treated equally, we should see somewhat of a worsening in 
health outcomes because the available resources are now distributed across more children.
If children are not treated equally, we should see much worse outcomes for the last
girl because the parents have an incentive to keep investing in the children they have
already rather than the new child.

How does education play into this?
The negative effects of shorter spacing is strongly mitigated by higher education.
Hence, women with more education are in the position that they can better tolerate short 
spacing in the absence of a boy and that the potential negative effects on the next child 
should be much less pronounced.
Furthermore, the lower mortality risk should show up whether the next child is a boy 
or a girl.
It is clearly still possible that there would some difference in mortality outcomes 
across boys and girls because of a larger investment in the boy once born, but it should
not be as large.

What are the effects of the introduction of sex selection?
The introduction of sex selection means that parents now have the option to avoid
birth of children with an undesired sex.
This impacts the observed birth spacing since, as discussed above, each sex-selective 
abortions increases the interval between births by six to twelve months.
The question is who is likely to use sex selection, when, and why?

[could I simply copy the results from my WB paper's theory section?]

There is clear evidence that in India the use of sex selection increases with education
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.
There are two potential reasons for this.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining stated son preference \citep{bhat03,pande07}.
}
First, women with more education generally have lower fertility either because of the
higher opportunity cost of time or because of other changes in the preferred number of
children.
Second, women with more education also tend to live in households with higher income,
which better enables them to access prenatal sex determination.
[It does not seem possible to cleanly separate out these differences]

should have something about how short spacing to conception could still be important.
Maybe even more so since health outcomes for short spacing birth likely has improved
over time.

@@

@@

@@
